---
title: "Farmland Biodiversity Health Index - <br> Broadleaf Arable Plants Risk Scores"
author: "Elin Falla"
output:
  html_document: default
  pdf_document: default
  word_document: default
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

This script calculates risk scores associated with changes to farm management, for **broadleaf arable plant** species reliant on farmland.  

**Note:** the script needs at least 2 species inputted to run.  

The method is from **Butler, S. J. et al.** 'A cross-taxonomic index for quantifying the health of farmland biodiversity.' J. Appl. Ecol. 46, 1154-1162 (2009) https://doi.org/10.1111/j.1365-2664.2009.01709.x  

This risk assessment framework assumes that the major mechanisms of impact on UK farmland boradleaf arable plants will be:  

1. reduced germination potential ($G_t$)   
2. reduced flowering potential ($F_t$)  
3. increased fertiliser application ($N_t$)
4. reduced soil moisture ($M_t$)

The method calculates risk based on the overlaps between the resource requirements used by each species and the resource requirements affected by the change.

&nbsp;

The risk score is calculated as follows: 

$$Risk Score = \frac{(G_t + F_t + N_t + M_t)} {R}$$

where $G_t$ is the risk score associated with reduced germination potential, $F_t$ is the risk score associated with reduced flowering potential, $N_t$ is the risk score associated with increased fertiliser application, $M_t$ is the risk score associated with reduced soil moisture and $R$ is the species’ reliance on farmland habitat.

&nbsp;

$G_t$ is calculated as follows:  
$$G_t = \frac{A}{G*H}$$

$A$ = number of points of coincidence between impact on and species' use of available germination periods
$G$ = number of germination periods used by the species
$H$ = number of habitat components used by the species 

-----  

$F_t$ is calculated as follows:  
$$F_t = \frac{B}{F*H}$$

$B$ = number of points of coincidence between impact on and species' use of available flowering periods
$G$ = number of flowering periods used by the species
$H$ = number of habitat components used by the species 

-----

$N_t$ is calculated as follows:  
$$N_t = \frac{C * (10 - E) + 1}{10}$$

$E$ = species' Ellenberg score for Nitrogen affinity
$C$ = number of points of coincidence between impact of increased nitrogen on habitats and species' use of those habitats

Note that this equation is different to Butler's original, but uses the same method he did. It just more clearly represents the process.  

-----

$M_t$ is calculated as follows:  
$$M_t = \frac{M}{10} * \frac{D}{H}$$

$M$ = species' Ellenberg score for soil moisture affinity  
$D$ = number of points of coincidence between impact of decreased moisture on habitats and species' use of those habitats  
$H$ = number of habitat components used by the species  

Note that this equation is different to Butler's original, but uses the same method he did. It just more clearly represents the process.  


## 2. Set-up
### 2.1. Load packages
```{r load-packages, message=FALSE, purl=FALSE}

rm(list = ls()) # clear env
library(readxl) # for reading in Excel sheets
library(tidyverse) # for data manipulation

```

## 2.2. Load necessary functions

```riskscore_functions.R``` contains functions used by all risk score scripts. It contains:  

* **check_aligned()** - It is essential to the method used here to ensure that the columns of the management change dataset and species' requirements dataset align before calculating overlap parameters - i.e. parameters that work out the impact of a change on a specific resource for a species. This function checks this, and outputs a warning if the column names of the two inputted datasets are not identical.  

* **NA_to_zero()** - Replaces all NAs in a dataset with zeros.  

``` {r check-alignment}
source("riskscore_functions.R")
```


### 2.3. Read in required datasets  

Two input files are required to calculate broadleaf arable plant risk scores:  

* Species' requirements - requirements for survival per species, entered as binary data (with the exception of moisture and nitrogen Ellenberg scores)  

* Farm management changes - binary data on how the farm management strategies affect the species' resources  

The files are read in and then all NA values are set to 0.  

Note that the farm management strategies are read in a separate chunk with ```purl = FALSE``` so if the file is sourced, a user can load their own management strategy dataset. It must however, follow the same format as the plant_man_strats dataset.  

``` {r read-in-data}
# plant requirements data
plants <- read_xlsx("Arable_plant_requirements.xlsx", skip = 1) 

# replace NAs with 0s 
plants <- NA_to_zero(plants)

```

``` {r load-man-strats, purl=FALSE}
# farm management strategies
plant_man_strats <- read_xlsx("change.xlsx", sheet="plants", skip = 1)

# replace NAs with 0s 
plant_man_strats <- NA_to_zero(plant_man_strats) 

```

## 3. Calculate $G_t$

$$G_t = \frac{A}{G*H}$$
$G_t$ is the risk score associated with reduced germination potential.  

### 3.1. Calculate $$ G * H $$ and $A$

Here, rather than separately calculating $H$ (number of habitat components used by each species) and $G$ (number of germination periods used by each species), I calculate $H * G$ (the denominator of the $G_t$ equation). I do this as the germination data is provided per habitat, so summing these data per species gives a calculation of $$ G * H $$.  

The calculation of $A$ uses a straightforward overlap between the management change effects on germination and species germination requirements. The management change is applied sequentially to each species (row of requirements data), and the '&' operator is used, meaning an impact is recorded if there is a non-zero value in both data sets in a particular column. This indicates both that the species uses the particular resource and that it's impacted by the management change. These impacts are then summed per species (row) to give a value for $A$ per species.

 
``` {r calc-Gt}
calc_plant_H_G_A <- function(data, strat) {
  
  # extract germination cols of species requirements and management strats
  germination <- data %>%
    select(germ_start : germ_end,
           -c(germ_start, germ_end))
  
    germination_effects <- strat %>%
    select(germ_start : germ_end,
           -c(germ_start, germ_end))
  
  # calculate H * G (denominator of Gt equation)
  H_times_G <- germination %>%
    rowSums(.)

  # check requirements and change data columns are aligned (i.e. the same)
  check_aligned(colnames(germination), colnames(germination_effects))
  
  # calculate overlap between germination requirements and effects
  overlap <- t(apply(germination, 1, function(row) row & germination_effects))  
  
  # calculate A
  A <- rowSums(overlap)
  
  return(data.frame(H_times_G, A))
}
```


### 3.2. Calculate $G_t$  

This function calls ```calc_plant_H_G_A()``` then uses the output to calculate $G_t$ for each species.  

``` {r calc-A}
calc_plant_Gt <- function(data, strat) {
  
  H_G_A <- calc_plant_H_G_A(data, strat)
  A <- H_G_A$A
  H_times_G <- H_G_A$H_times_G
  
  # CALCULATE Gt
  Gt <- A / H_times_G
  
  #replace infinite or NA values with 0
  Gt <- 
    ifelse(
      is.nan(Gt) | is.infinite(Gt) == T,  # test
      0, # if yes
      Gt  # if no
    ) 
  
  return(Gt)
}
```


## 4. Calculate $F_t$

$$F_t = \frac{B}{F*H}$$

$F_t$ is the risk score associated with reduced flowering potential.  

### 4.1. Calculate $$ F * H $$ and $B$

Here, similar to in the calculation of $G_t$, I calculate $H * F$ (denominator of $F_t$ equation) directly, rather than calculating $H$ and $F$ separately. This was due to the fact that the flowering data was inputted per habitat, as for in the calculation of $G_t$.  

$B$ is also calculated in the same way as $A$ is in the calculation of $G_t$, but using the flowering data from species' requirements and farm management datasets, rather than germination data - see explanation of $A$ above.  

``` {r calc-F-H_B}
calc_plant_H_F_B <- function(data, strat) {
  
  # extract flowering data from species requirements and management strat
  flowering <- data %>%
    select(flower_start : flower_end,
           -c(flower_start, flower_end))
  
  flowering_effects <- strat %>%
  select(flower_start : flower_end,
         -c(flower_start, flower_end))
  
  # calculate F * H (denominator of Ft equation)
  F_times_H <- flowering %>%
    rowSums(.)
  
  # check requirements and change data columns are aligned (i.e. the same)
  check_aligned(colnames(flowering), colnames(flowering_effects))
  
  # calculate overlap between flowering requirements and change effects
  overlap <- t(apply(flowering, 1, function(row) row & flowering_effects))  
  
  # calculate B
  B <- rowSums(overlap)
  
  return(data.frame(F_times_H, B))
}

```

``` {r test-F-H-B, purl=FALSE}
calc_plant_H_F_B(plants, plant_man_strats[1,])
```

### 4.2. Calculate $F_t$

This function calc_plants $F_t$ for each species by calling ```calc_plant_H_F_B()```.  

``` {r calc-Ft}
calc_plant_Ft <- function(data, strat) {
  
  H_F_B <- calc_plant_H_F_B(data, strat)
  B <- H_F_B$B
  F_times_H <- H_F_B$F_times_H
  
  # CALCULATE Gt
  Ft <- B / F_times_H

  #replace infinite or NA values with 0
  Ft <- 
    ifelse(
      is.nan(Ft) | is.infinite(Ft) == T,  # test
      0, # if yes
      Ft  # if no
    ) 

  return(Ft)
}
```


## 5. Calculate $N_t$ and $M_t$

$$N_t = \frac{C * (10 - E) + 1}{10}$$
$N_t$ is the risk score associated with increased fertiliser application.  

$$M_t = \frac{M}{10} * \frac{D}{H}$$

$M_t$ is the risk score associated with reduced soil moisture.  

----- 

For the calculation of $N_t$ and $M_t$, $C$ and $D$ are their respective overlap parameters, recording the impact of the change on habitats and species' use of those habitats. This is calculated in the same way as $A$ and $B$ - see explanation of $A$ for more detail.

In the calculation of $N_t$, $$\frac{(10 - E + 1)}{10}$$ is the base calculation of the potential impact of increased nitrogen on a particular species, based on its Ellenberg nitrogen score ($E$). For $M_t$, the same is true for $$\frac{M}{10}$$. Whether this potential impact is seen depends on the respective overlap parameter ($C$ and $D$). For $M_t$, the risk is calculated per habitat (hence the division by $H$) and summed, leading to a maximum risk of $$\frac{M}{10}$$ if all habitats are affected. For $N_t$ however, this is not the case; the potential total risk is 3 times $$\frac{(10 - E + 1)}{10}$$ (as there are 3 habitats), which would correspond to a $C$ value of 3, indicating all habitats are affected. The reason for this difference is that it is very unlikely that increased nitrogen via fertiliser application will be relevant to habitats other than the cropped area, whereas the effect of reduced moisture is more likely to be seen across multiple habitats, making a per-habitat calculation necessary.    

``` {r calc-Nt-Mt}
calc_plant_Nt_Mt <- function(data, strat) {
  
  # select nitrogen affinity cols from plant requirements and man_strats
  nitrogen <- data %>%
    select(nitro_start : nitro_end,
           -c(nitro_start, nitro_end))
  
  nitrogen_effects <- strat %>%
  select(nitro_start : nitro_end,
         -c(nitro_start, nitro_end))

  # select moisture affinity cols from plant requirements and man_strats
  moisture <- data %>%
    select(moist_start : moist_end,
           -c(moist_start, moist_end))
  
  moisture_effects <- strat %>%
    select(moist_start : moist_end,
           -c(moist_start, moist_end))

  # extract E: Ellenberg nitrogen affinity score
  E <- data %>% 
    pull(Nitrogen)
  
  # extract M: Ellenberg moisture affinity score
  M <- data %>% 
    pull(Moisture)

  # calculate H: number of habitats each species occupies (for calculation of Mt)
  H <- data %>%
    select(hab_start : hab_end) %>%
    rowSums(.)

  # check column alignment for nitrogen and moisture data
  check_aligned(colnames(nitrogen), colnames(nitrogen_effects))
  check_aligned(colnames(moisture), colnames(moisture_effects))
  
  # calculate C and D: overlaps between requirements and management strat for moisture and nitrogen
  C_temp <- t(apply(nitrogen, 1, function(row) row & nitrogen_effects)) 
  D_temp <- t(apply(moisture, 1, function(row) row & moisture_effects))
  
  C <- rowSums(C_temp)
  D <- rowSums(D_temp)
  
  # CALCULATE NT AND MT
  Nt <- C * (10 - E + 1) / 10
  Mt <- D/H * M/10
  
  return(data.frame(Nt, Mt))
}
```

``` {r test-Nt_mtm, purl=FALSE}
# plant_validation <- read_xls("Butler_validation/Validation of weed matrix.xls", 3)
# 
# dat <- data.frame(species = all_species,
#                   Nt = calc_plant_Nt_Mt(plants, plant_man_strats[4,]))
# dat[dat$Nt != 0,]
# dat[80,]
# calc_plant_Nt_Mt(plants, plant_man_strats[2,])



# for (row in 1:nrow(plants)) {
#   if (is.na(plants[row, "Species2"]) == T) {
#     break
#   }
#   
#   index <- which(plants$Species == as.character(plants[row, "Species2"]))
#   plants[index, "Nitrogen"] <- as.numeric(plants[row, "Ellenberg"])
# }
# 
# write_xlsx(plants, "plants_nitrogen.xlsx")

```


## 6. Calculate risk scores

### 6.1 Make function for autumn sowing condition (one of Butler's management changes)
The 'autumn sowing' management change used by Butler causes decline for July flowering crop habitat plants **only if** they are autumn germinating.  

I've therefore created a function that allows the addition of a condition to a parameter after calculating it. In this specific case, implemented in the ```calc_plant_riskscore()``` function below, I'm adding a condition that the species must be autumn germinating to $F_t$ (the effects on flowering).  

``` {r aut-sow}
add_condition <- function(data, parameter, condition) {
  
  # condition will be non-zero if species uses resources in any of conditioned cols
  condition <- data %>%
    select(contains(condition)) %>%
    rowSums(.)
  
  # if condition is non-zero, leave parameter, else set parameter to 0.
  conditioned_param <- 
    ifelse(
      condition != 0,
      parameter,
      0
    )
  
  return(conditioned_param)
}

```

### 6.2. Calculate risk scores

The below function calls all the functions defined above in order to calculate a final risk score. Note that, for the functions defined up to this point, parameters are calculated for a single management change. The function below can handle multiple management changes, and will output a risk score for each, as well as a total risk score, calculated by summing up the risk scores of all the changes.  

``` {r calc-risk}
calc_plant_riskscore <- function(data, strategies) {
  
  # initialise risk scores dataframe, rows are species, cols are strategies
  riskscores <- data.frame(matrix(NA, 
                                  nrow = nrow(data), 
                                  ncol = nrow(strategies) + 1)) # +1 to add in species names as first col
  # put species names in first col
  riskscores[,1] <- data[,1]
  
  # name columns of risk scores according to strategies
  strat_names <- strategies %>% pull(agric_change)
  colnames(riskscores) <- c("species", strat_names)
  
  # extract reliance (R) values from plant requirements data
  R <- data[,"Reliance"]  
  
  # CALCULATE RISK SCORES per strategy and store in risk scores dataframe
  for (strat in 1:nrow(strategies)) {
    
    Gt <- calc_plant_Gt(data, strategies[strat,])
    Ft <- calc_plant_Ft(data, strategies[strat,])

    if (strat_names[strat] == "aut_sow") { # for autumn sowing Butler change
      Ft <- add_condition(data, Ft, "_Au")  # add the condition that the species germinates in autumn to Ft
    }
    
    Nt_Mt <- calc_plant_Nt_Mt(data, strategies[strat,])
    Nt <- Nt_Mt$Nt
    Mt <- Nt_Mt$Mt
    
    # CALCULATE RISK SCORE
    riskscores[,strat+1] <- (Gt + Ft + Nt + Mt) / R
    
  }
  
  # calculate total risk for each species by summing risk across all strategies (if more than one strategy)
  if (nrow(strategies) > 1) {
    riskscores$total_risk <- rowSums(riskscores[,-1])
  }
  
  return(riskscores)
}
```

### 6.3. Calculate standardised risk score

In order to allow cross-taxonomic comparison of risk scores, they must be standardised. This is done by dividing the risk scores by the number of key resource requirements in the framework. This is equivalent to the number of overlap parameters - i.e. parameters that work out the impact of a change on a specific resource for a species. For broadleaf arable plants, this number is 4, as there are 4 such parameters: $A$, $B$, $C$ and $D$. See equations in Section 1 for more details.  

If only comparing risk scores within the taxon, this standardisation step is not required, and the ```calc_plant_riskscore()``` function can be used instead.  

``` {r calc-standardised-riskscore}
calc_std_plant_riskscore <- function(data, strategies) {
  
  # set number of resource requirements used by the taxa
  num_resource_reqs <- 4
  
  # calculate non-standardised risk scores
  risk_scores <- calc_plant_riskscore(data, strategies)

  # divide all risk scores by number of resource requirements (minus the species name column)
  std_risk_scores <- risk_scores[,-1] / num_resource_reqs
  std_risk_scores <- cbind(risk_scores$species, std_risk_scores) # add back species name column
  names(std_risk_scores)[1] <- "species"
  
  return(std_risk_scores)
}
```


### 6.4. Test risk scores

Calculate Butler's risk scores to they are the same as his.  

``` {r test-risk, purl=FALSE}
risk <- calc_plant_riskscore(plants, plant_man_strats)
#risk[which(butler$same == F),]
#butler[which(butler$same == F),]

butler <- read_xlsx("butler_plants_risk.xlsx")

for (sp in butler$species) {
  if (!(sp %in% risk$species)) {
    print(sp)
    butler <- butler[butler$species != sp,]
  }
}

butler$my_risk <- round(risk$total_risk, 2)
butler$same <- butler$my_risk == butler$total_risk
butler$diff <- butler$my_risk - butler$total_risk
butler
print(paste("correct species =", sum(butler$same), "/", nrow(risk))) # 26. 28 with Nt/Mt counted once

which(butler$my_risk != butler$total_risk * -1)
```