---
title: "Farmland Biodiversity Health Index - <br> Broadleaf Arable Plants Risk Scores"
author: "Elin Falla"
output:
  html_document: default
  pdf_document: default
  word_document: default
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

This script calculates risk scores associated with changes to farm management, for **broadleaf arable plant** species reliant on farmland.  

**Note:** the script needs at least 2 species inputted to run.  

The method is from **Butler, S. J. et al.** 'A cross-taxonomic index for quantifying the health of farmland biodiversity.' J. Appl. Ecol. 46, 1154-1162 (2009) https://doi.org/10.1111/j.1365-2664.2009.01709.x  

This risk assessment framework assumes that the major mechanisms of impact on UK farmland boradleaf arable plants will be:  

1. reduced germination potential ($G_t$)   
2. reduced flowering potential ($F_t$)  
3. increased fertiliser application ($N_t$)
4. reduced soil moisture ($M_t$)

The method calculates risk based on the overlaps between the resource requirements used by each species and the resource requirements affected by the change.

&nbsp;

The risk score is calculated as follows: 

$$Risk Score = \frac{(G_t + F_t + N_t + M_t)} {R}$$  

where $G_t$ is the risk score associated with reduced germination potential, $F_t$ is the risk score associated with reduced flowering potential, $N_t$ is the risk score associated with increased fertiliser application, $M_t$ is the risk score associated with reduced soil moisture and $R$ is the species’ reliance on farmland habitat.

&nbsp;

$G_t$ is calculated as follows:  
$$G_t = \frac{A}{G*H}$$

$A$ = number of points of coincidence between impact on and species' use of available germination periods
$G$ = number of germination periods used by the species
$H$ = number of habitat components used by the species 

-----  

$F_t$ is calculated as follows:  
$$F_t = \frac{B}{F*H}$$  

$B$ = number of points of coincidence between impact on and species' use of available flowering periods
$G$ = number of flowering periods used by the species
$H$ = number of habitat components used by the species 

-----

$N_t$ is calculated as follows:  
$$N_t = \frac{(10 - E) + 1}{10}$$  

$E$ = Ellenburg score for Nitrogen affinity

-----

$M_t$ is calculated as follows:  
$$M_t = \frac{M}{10}$$

$M$ = Ellenburg score for soil moisture affinity

## 2. Set-up
### 2.1. Load packages
```{r load-packages, message=FALSE}

rm(list = ls()) # clear env
library(readxl) # for reading in Excel sheets
library(tidyverse) # for data manipulation

```


### 2.2. Read in required datasets  
Load datasets and replace all NAs with 0s.  
``` {r read-in-data}

# farm management strategies
man_strats <- read_xlsx("change.xlsx", 6, skip = 1)

# mammal matrix data
plants <- read_xlsx("Arable_plant_requirements.xlsx", skip = 1) 


# replace NAs with 0s in both datasets
man_strats <- man_strats %>% 
  mutate_all(~replace(., is.na(.), 0))

plants <- plants %>% 
  mutate_all(~replace(., is.na(.), 0))

```

## 3. Calculate $G_t$

### 3.1. Calculate $$ H * G $$ and $A$

Here rather than separately calculating $H$ and $G$, I calculate $H * G$ (the denominator of the $G_t$ equation) as this is easily calculated from the germination data, as it's written in as germination periods per habitat.  

``` {r calc-Gt}
calculate_H_G_A <- function(data, strat) {
  
  germination <- data %>%
    select(germ_start : germ_end,
           -c(germ_start, germ_end))
  
  # calculate H * G (denominator of Gt equation)
  H_times_G <- germination %>%
    rowSums(.)
  
  # calculate A
  germination_effects <- strat %>%
    select(germ_start : germ_end,
           -c(germ_start, germ_end))
  
  overlap <- t(apply(germination, 1, function(row) row & germination_effects))  
  
  A <- rowSums(overlap)
  
  return(data.frame(H_times_G, A))
}
```

``` {r test-H-G-A}
calculate_H_G_A(plants, man_strats[1,])
```

### 3.2. Calculate $G_t$  

``` {r calc-A}
calculate_Gt <- function(data, strat) {
  H_G_A <- calculate_H_G_A(data, strat)
  A <- H_G_A$A
  H_times_G <- H_G_A$H_times_G
  
  # CALCULATE Gt
  Gt <- A / H_times_G
  
  #replace infinite or NA values with 0
  Gt <- 
    ifelse(
      is.nan(Gt) | is.infinite(Gt) == T,  # test
      0, # if yes
      Gt  # if no
    ) 
  
  return(Gt)
}
```

## 4. Calculate $F_t$

### 4.1. Calculate $$ F * H $$ and $B$

Here, similar to the calculation of $G_t$, I calculate $$H * F$$ (denominator of $F_t$ equation) directly, rather than calculating $H$ and $F$ separately.  

``` {r calc-F-H_B}
calculate_H_F_B <- function(data, strat) {
  
  flowering <- data %>%
    select(flower_start : flower_end,
           -c(flower_start, flower_end))
  
  # calculate F * H (denominator of Ft equation)
  F_times_H <- flowering %>%
    rowSums(.)
  
  # calculate B
  flowering_effects <- strat %>%
    select(flower_start : flower_end,
           -c(flower_start, flower_end))
  
  overlap <- t(apply(flowering, 1, function(row) row & flowering_effects))  
  
  B <- rowSums(overlap)
  
  return(data.frame(F_times_H, B))
}

```

``` {r test-F-H-B}
calculate_H_F_B(plants, man_strats[1,])
```

### 4.2. Calculate $F_t$
``` {r calc-Ft}
calculate_Ft <- function(data, strat) {
  H_F_B <- calculate_H_F_B(data, strat)
  B <- H_F_B$B
  F_times_H <- H_F_B$F_times_H
  
  # CALCULATE Gt
  Ft <- B / F_times_H

  #replace infinite or NA values with 0
  Ft <- 
    ifelse(
      is.nan(Ft) | is.infinite(Ft) == T,  # test
      0, # if yes
      Ft  # if no
    ) 

  return(Ft)
}
```

``` {r }
calculate_Ft(plants, man_strats[1,])
```


## Calculate $N_t$ and $M_t$
``` {r calc-Nt-Mt}
calculate_Nt_Mt <- function(data, strat) {
  
  # select nitrogen affinity cols from plant requirements and man_strats
  nitrogen <- data %>%
    select(nitro_start : nitro_end,
           -c(nitro_start, nitro_end))
  
  nitrogen_effects <- strat %>%
    select(nitro_start : nitro_end,
           -c(nitro_start, nitro_end))
  
  # select moisture affinity cols from plant requirements and man_strats
  moisture <- data %>%
    select(moist_start : moist_end,
           -c(moist_start, moist_end))
  
  moisture_effects <- strat %>%
    select(moist_start : moist_end,
           -c(moist_start, moist_end))
  
  # extract number of habitats each species occupies (to work out E and M)
  num_habs <- data %>%
    select(hab_start : hab_end) %>%
    rowSums(.)
  
  # calculate E for each species (species nitrogen affinity)
  E <- nitrogen %>%
    transmute(E = rowSums(.) / num_habs) 
  
  # calculate M for each species (species moisture affinity)
  M <- moisture %>%
    transmute(M = rowSums(.) / num_habs)
  
  # calculate overlaps between requirements and management strat for moisture and nitrogen
  nitro_overlap <- t(apply(nitrogen, 1, function(row) row & nitrogen_effects)) 
  moist_overlap <- t(apply(moisture, 1, function(row) row & moisture_effects))
  
  # calculate E and M taking management strat into account
  E_effects <- rowSums(nitro_overlap) * E
  M_effects <- rowSums(moist_overlap) * M
  
  # convert E_effects to vector to use ifelse function
  E_effects <- pull(E_effects, E)
    M_effects <- pull(M_effects, M)
      E <- pull(E, E)
      M <- pull(M, M)


  # # calculate Nt
  # Nt <-
  #   ifelse(
  #     E_effects == 0,  # test
  #     0,  # if yes
  #     (10 - E_effects + 1) / 10  # if no
  #   )
  # 
  # # calculate Mt
  # Mt <- M_effects / 10

  Nt <-
  ifelse(
    E_effects == 0,  # test
    0,  # if yes
    (10 - E + 1) / 10  # if no
  )

  Mt <-
  ifelse(
    M_effects == 0,  # test
    0,  # if yes
    M / 10  # if no
  )
  
  output <- data.frame(Nt, Mt)
  names(output) <- c("Nt", "Mt")
  
  return(output)
}
```

``` {r test-Nt_mt}
calculate_Nt_Mt(plants, man_strats[2,])[36,]
plants[36,]
```

## Calculate risk scores

### Make function for autumn sowing condition
Autumn sowing management change causes decline for July flowering crop habitat plants **only if** they are autumn germinating.  

I've therefore created a function that allows the addition of a condition to a parameter after calculating it. In this specific case, I'm adding a condition that the species must be autumn germinating to $F_t$, which is implemented in calc_plant_risk().  

``` {r aut-sow}
add_condition <- function(data, parameter, condition) {
  
  condition <- data %>%
    select(contains(condition)) %>%
    rowSums(.)
  
  conditioned_param <- 
    ifelse(
      condition != 0,
      parameter,
      0
    )
  
  return(conditioned_param)
}

```



``` {r calc-risk}
calc_plant_risk <- function(data, strategies) {
  
  # initialise risk scores dataframe, rows are species, cols are strategies
  riskscores <- data.frame(matrix(NA, 
                                  nrow = nrow(data), 
                                  ncol = nrow(strategies) + 1)) # +1 to add in species names as first col
  # put species names in first col
  riskscores[,1] <- data[,1]
  
  # name columns of risk scores according to strategies
  strat_names <- strategies %>% pull(agric_change)
  colnames(riskscores) <- c("species", strat_names)
  
  # extract reliance (R) values from plant requirements data
  R <- data[,"Reliance"]  
  
  # CALCULATE RISK SCORES per strategy and store in risk scores dataframe
  for (strat in 1:nrow(strategies)) {
    
    Gt <- calculate_Gt(data, strategies[strat,])
    Ft <- calculate_Ft(data, strategies[strat,])

    if (strat_names[strat] == "aut_sow") { # for autumn sowing change
      Ft <- add_condition(data, Ft, "_Au")  # add the condition that the species germinates in autumn to Ft
    }
    
    Nt_Mt <- calculate_Nt_Mt(data, strategies[strat,])
    Nt <- Nt_Mt$Nt
    Mt <- Nt_Mt$Mt
    
    riskscores[,strat+1] <- (Gt + Ft + Nt + Mt) / R
    
  }
  
  # calculate total risk for each species by summing risk across all strategies (if more than one strategy)
  if (nrow(strategies) > 1) {
    riskscores$total_risk <- rowSums(riskscores[,-1])
  }
  
  return(riskscores)
}
```

``` {r test-risk}
risk <- calc_plant_risk(plants, man_strats)
butler <- read_xlsx("butler_plants_risk.xlsx")

for (sp in butler$species) {
  if (!(sp %in% risk$species)) {
    print(sp)
    butler <- butler[butler$species != sp,]
  }
}

butler$my_risk <- round(risk$total_risk, 2)
butler$same <- butler$my_risk == butler$total_risk
butler$diff <- butler$my_risk - butler$total_risk
butler
print(paste("correct species =", sum(butler$same), "/", nrow(risk))) # 26. 28 with Nt/Mt counted once

plants[butler$same == T,]

nrow(plants[plants$C_hab == 1,]) # 102
nrow(plants[plants$H_Herbi == 0 & plants$M_Herbi == 0 & plants$C_Herbi == 0 ,]) # 85
nrow(plants[plants$H_Moisture == 2 | plants$M_Moisture == 2 | plants$C_Moisture == 2 ,]) # 173

before <- risk[36,]
before
after <- risk[36,]
after
```