---
title: "Farmland Biodiversity Health Index - <br> Mammal Risk Scores"
author: "Elin Falla"
output:
  html_document: default
  pdf_document: default
  word_document: default
fontsize: 11pt
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction  
This script calculates risk scores associated with changes to farm management, for **butterfly** species reliant on farmland.  

**Note:** the script needs at least 2 species inputted to run.

The method is from **Butler, S. J. et al.** 'A cross-taxonomic index for quantifying the health of farmland biodiversity.' J. Appl. Ecol. 46, 1154-1162 (2009) https://doi.org/10.1111/j.1365-2664.2009.01709.x  

## Set-up
### Load packages
```{r load-packages, message=FALSE}

rm(list = ls()) # clear env
library(readxl) # for reading in Excel sheets
library(tidyverse) # for data manipulation

```

### Read in required datasets  
Load datasets and replace all NAs with 0s.  
``` {r read-in-data}

# farm management strategies
man_strats <- read_xlsx("change.xlsx", 5) 

# butterfly matrix data
butterfly <- read_xlsx("Butterfly matrix.xlsx", skip = 3) 

# forage plant locations
plant_locs <- read_xlsx("forage_plant_location.xlsx", na="NA")
plant_locs <- plant_locs[,-6]  # remove 'cropped area' column as not needed
plant_locs <- plant_locs[is.na(plant_locs$riparian) == F,]  # remove NA rows, as this means the species isn't relevant to butterflies (bees only)

# replace NAs with 0s in both datasets
man_strats <- man_strats %>% 
  mutate_all(~replace(., is.na(.), 0))

butterfly[is.na(butterfly)] <- 0

plant_locs <- plant_locs %>% 
  mutate_all(~replace(., is.na(.), 0))
```

### Write function to create separate adult and larval datasets
``` {r sep-a-l}
separate_lifestages <- function(data) {
  
  # create separate dataframes for adults and larvae
  adults <- data %>%
    select(contains("Ad_"))
  colnames(adults) <- gsub(pattern = "Ad_", replacement = "", colnames(adults))  # rename cols to not include "Ad_"

  larvae <- data %>%
    select(contains("L_"))
  colnames(larvae) <- gsub(pattern = "L_", replacement = "", colnames(larvae))  # rename cols to not include "L_"
 
  return(list(adults, larvae))
}
```
## Calculate $P_t$ (for adults, $P_{ta}$, and larvae, $P_{tl}$)

### Calculate $G$ and $H$
$G$ = number of generations across all activity periods (ie. number of generations per year)
$H$ = number of habitat components used by the species 


how many diff gens of adults/larvae in a year - just the generations column in butt matrix

``` {r calc-Ga-Gl}
calculate_Ga_Gl_Ha_Hl <- function(data) {

  #separate the adult and larval lifestages into separate dataframes (in order to loop over them)
  lifestages <- separate_lifestages(data)
  
  # create output dataframe
  output <- data.frame(Ha = rep(NA, nrow(data)),
                       Hl = rep(NA, nrow(data)),
                       Ga = rep(NA, nrow(data)),
                       Gl = rep(NA, nrow(data)))
  
  # calculate G and H for both adults and larvae
  for (i in 1:length(lifestages)) {
    
    # calculate H
    output[,i] <- lifestages[[i]] %>%
      select(H_hab : R_hab) %>%
      rowSums(.)
 
    # calculate G (select generations col of respective df)
    output[,i+2] <- lifestages[[i]] %>%
      select(generations)
  }
    
  return(output)
}
```

test G + H func:
``` {r test-GH}
calculate_Ga_Gl_Ha_Hl(butterfly)

```

### Calculate $G_t$
$G_t$ = number of generations of a species active in the activity periods affected

Here, 'strat' is a single management strategy - i.e. one row of the management strategy dataframe.

number of each gens foraging habitats affected
``` {r calc-Gta-Gtl}
calculate_Gta_Gtl <- function(data, strat) {
  
  # create separate adult and larvae dataframes
  lifestages <- separate_lifestages(data)
  
  # select relevant cols of strat
  activity_effects <- strat %>%
      select(H_early : R_late)
  
  # create output dataframe
  output <- data.frame(Gta = rep(NA, nrow(data)),
                       Gtl = rep(NA, nrow(data)))
  
  # calculate Gt for bot h adults and larvae
  for (i in 1:length(lifestages)) {
    
    activity <- lifestages[[i]] %>%
      select(H_early : R_late)
  
  # Check columns of diets_habs and strat are the same (if not aligned, overlap calculation will not be correct)
  aligned <- identical(colnames(activity), colnames(activity_effects))
  
  if (aligned == F) {
    warning(paste0("WARNING: Butterfly data and management strategy columns do not align/are not the same - Gt calculations will be incorrect. Iteration = ", i, ", 1=adult, 2=larvae"))
  }
  
  # find overlap
  overlap <- t(apply(activity, 1, function(row) row & activity_effects)) # NOTE: THIS ASSUMES THE COLUMNS ALIGN
  
  # sum rows to find Gt and store in output dataframe
  output[,i] <- rowSums(overlap)
  
  }
  
  return(output)
  
}
```

test calc Gt:
``` {r test-Gt}
calculate_Gta_Gtl(butterfly, man_strats[5,])[12,]
```

### Calculate Pt
Pta = Gta/(Ga*Ha)
Ptl = Gtl/(Gl*Hl)

``` {r calc-Pt}
calculate_Pta_Ptl <- function(data, strat) {
  
  G_H <- calculate_Ga_Gl_Ha_Hl(data)
  Ga <- G_H$Ga
  Gl <- G_H$Gl
  Ha <- G_H$Ha
  Hl <- G_H$Hl
  
  Gt <- calculate_Gta_Gtl(data, strat)
  Gta <- Gt$Gta
  Gtl <- Gt$Gtl
  
  # CALCULATE Pta
  Pta = Gta / (Ga * Ha)
  
  # replace infinite or NA values with 0
  Pta <- 
  ifelse(
    is.nan(Pta) | is.infinite(Pta) == T,  # test
    0, # if yes
    Pta  # if no
  )

  # CALCULATE Ptl
  Ptl = Gtl / (Gl * Hl)
  
  # replace infinite or NA values with 0
  Ptl <- 
  ifelse(
    is.nan(Ptl) | is.infinite(Ptl) == T,  # test
    0, # if yes
    Ptl  # if no
  )
  
  return(data.frame(Pta, Ptl))
}
```

``` {r or-cols-func}
apply_or_to_cols <- function(data) {
  
  # if data isn't a dataframe, return it
  if (is.data.frame(data) == F) {
    return(data)
  }
  # if data has only 1 column, return it
  if (ncol(data) <= 1) {
    return(data)
  }
  
  result <- data[,1]  # set first column as current column
  
  # use or operator across all columns to find all non-zero instances, put into a single col
    for (i in 2:ncol(data)) {
      result <- result | data[,i]
    }
  
  return(result)
  
}
```

## Calculate Ft (for a and l)
### Calculate A
$A$ = number of points of coincidence between the impact on and species’ use of forage plant families
### Calculate F
$F$ = points of coincidence between species habitat use and the location of its forage plants

Butterflies have long tongues, so adults can forage on any plant in the habitat(s) they occupy. Therefore, Fa (F for adults) will just be the total number of forage plants that are found in the species habitat(s).

for adults: SUBSET OUT POACAE

FOR LARVAE: need to see if diff poacae species grow in different habitats or not
``` {r calc-Fa-Aa}
calculate_Fa_Aa <- function(data, plant_locs, strat) {
  
  # PUT IN CHECK THAT COLUMNS ALIGN
  
  # remove Poaceae (grasses) from plant species in plant_locs as adults don't forage on them
  plant_locs <- plant_locs[plant_locs$plantfamily != "Poaceae",]
  
  # remove species name col of plant locs (so columns align with habs)
  plant_locs <- plant_locs[,-1]
  
  # subset butterfly data to habitat columns
  habs <- data %>%
    select(Ad_H_hab : Ad_R_hab, )

  # select hab cols of change data
  hab_effects <- strat %>%
    select(H_hab : R_hab)
  damp_effects <- strat %>% 
    select(H_damp : R_damp)

  effects <- list(hab_effects, damp_effects)
  plants_affected <- data.frame(matrix(ncol = length(effects), nrow =nrow(hab_effects)))
  
  for (i in 1:length(effects)) {
      overlap <- apply(habs, 1, function(row) row & effects[[i]])
    for(species in 1:nrow(overlap)) {
      
      species <- overlap[species,]
      
      plant_locs_affected <- plant_locs[,which(species != 0)]
      if (ncol(plant_locs_affected) == 0) {
        plant_locs_affected <- 0
      }
      
      plants_affected[,species] <- apply_or_to_cols(plant_locs_affected)
      
      if (i == 2) {  # if damp_effects
        plants_affected <- plants_affected & plant_locs$likesdamp == "yes"
      }
      
      
      
    }
    
  }
  #total_plants_affected <- 

  
  # subset plant_locs to include only habitats affected by the strategy
  locs_affected <- plant_locs[,which(hab_effects != 0)]
  if (ncol(locs_affected) == 0) {
    locs_affected <- 0
  }
  print("locs_affected:")
  print(locs_affected)
  print("end locs_affected")
  
  # find all plants affected by applying the or operator to all habitats affected
  locs_plants_affected <- apply_or_to_cols(locs_affected)
  
  # do same for damp plants affected:
  # subset affected habitats...
  damp_affected <- plant_locs[,which(damp_effects != 0)]
  if (ncol(damp_affected) == 0) {
    damp_affected <- 0
  }
  # ...and use or operator to find all plants in those habitats
  all_damp_affected <- apply_or_to_cols(damp_affected)
  
  # reduce this list to only plants with a preference for damp
  damp_plants_affected <- all_damp_affected & plant_locs$likesdamp == "yes"
  
  # use or operator on plants and damp plants affected to get all plants affected  
  all_plants_affected <- locs_plants_affected | damp_plants_affected
  
  # initialise output arrays
  Fa <- 0
  Aa <- 0
  
  # calculate Fa and Aa for every species (ie. every row in habs dataframe)
  for (i in 1:nrow(habs)) {
    
    locs <- plant_locs[,which(habs[i,] == 1)] # extract cols of plant_locs for which butterfly habs=1
    
    all_plants <- apply_or_to_cols(locs)
      
    # Fa is total number of plants = sum the result
    Fa[i] <- sum(all_plants)
    
    # Aa is the sum of the overlap between the plants used by the species (all_plants) and those          # affected by the change (all_plants_affected)
    
    Aa[i] <- sum(all_plants & all_plants_affected)
   
  }
  return(data.frame(Fa, Aa))
}
```

``` {r test-Fa}
calculate_Fa_Aa(butterfly, plant_locs, man_strats[1,])
```

