---
title: "Farmland Biodiversity Health Index - <br> Mammal Risk Scores"
author: "Elin Falla"
output:
  html_document: default
  pdf_document: default
  word_document: default
fontsize: 11pt
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction  
This script calculates risk scores associated with changes to farm management, for **butterfly** species reliant on farmland.  

**Note:** the script needs at least 2 species inputted to run.

The method is from **Butler, S. J. et al.** 'A cross-taxonomic index for quantifying the health of farmland biodiversity.' J. Appl. Ecol. 46, 1154-1162 (2009) https://doi.org/10.1111/j.1365-2664.2009.01709.x  

## Set-up
### Load packages
```{r load-packages, message=FALSE}

rm(list = ls()) # clear env
library(readxl) # for reading in Excel sheets
library(tidyverse) # for data manipulation

```

### Read in required datasets  
Load datasets and replace all NAs with 0s.  
``` {r read-in-data}

# farm management strategies
man_strats <- read_xlsx("change.xlsx", 5) 

# butterfly matrix data
butterfly <- read_xlsx("Butterfly matrix.xlsx", skip = 3) 

# forage plant locations
plant_locs <- read_xlsx("forage_plant_location.xlsx")

# replace NAs with 0s in both datasets
man_strats <- man_strats %>% 
  mutate_all(~replace(., is.na(.), 0))

butterfly[is.na(butterfly)] <- 0
```

### Write function to create separate adult and larval datasets
``` {r sep-a-l}
separate_lifestages <- function(data) {
  
  # create separate dataframes for adults and larvae
  adults <- data %>%
    select(contains("Ad_"))
  colnames(adults) <- gsub(pattern = "Ad_", replacement = "", colnames(adults))  # rename cols to not include "Ad_"

  larvae <- data %>%
    select(contains("L_"))
  colnames(larvae) <- gsub(pattern = "L_", replacement = "", colnames(larvae))  # rename cols to not include "L_"
 
  return(list(adults, larvae))
}
```
## Calculate $P_t$ (for adults, $P_{ta}$, and larvae, $P_{tl}$)

### Calculate $G$ and $H$
$G$ = total number of life cycle components (i.e. sum of the number of generations in each activity period)
$H$ = number of habitat components used by the species 

``` {r calc-Ga-Gl}
calculate_Ga_Gl_Ha_Hl <- function(data) {
  generations <- 1 # FOR NOW

  #separate the adult and larval lifestages into separate dataframes (in order to loop over them)
  lifestages <- separate_lifestages(data)
  
  # create output dataframe
  output <- data.frame(Ha = rep(NA, nrow(data)),
                       Hl = rep(NA, nrow(data)),
                       Ga = rep(NA, nrow(data)),
                       Gl = rep(NA, nrow(data)))
  
  # calculate G and H for both adults and larvae
  for (i in 1:length(lifestages)) {
    
    # calculate H
    output[,i] <- lifestages[[i]] %>%
      select(H_hab : R_hab) %>%
      rowSums(.)
 
    # calculate G
    G_temp <- lifestages[[i]] %>%
      select(H_early : R_late) %>%
      rowSums(.)
    
    output[,i+2] <- (G_temp / output[,i]) * generations
  }
    
  return(output)
}
```

test G + H func:
``` {r test-GH}
calculate_Ga_Gl_Ha_Hl(butterfly)

```

### Calculate $G_t$
$G_t$ = number of generations of a species active in the activity periods affected

Here, 'strat' is a single management strategy - i.e. one row of the management strategy dataframe.
``` {r calc-Gta-Gtl}
calculate_Gta_Gtl <- function(data, strat) {
  
  generations <- 1 # FOR NOW
  
  # create separate adult and larvae dataframes
  lifestages <- separate_lifestages(data)
  
  # select relevant cols of strat
  activity_effects <- strat %>%
      select(H_early : R_late)
  
  # create output dataframe
  output <- data.frame(Gta = rep(NA, nrow(data)),
                       Gtl = rep(NA, nrow(data)))
  
  # calculate Gt for bot h adults and larvae
  for (i in 1:length(lifestages)) {
    
    activity <- lifestages[[i]] %>%
      select(H_early : R_late)
  
  # Check columns of diets_habs and strat are the same (if not aligned, overlap calculation will not be correct)
  aligned <- identical(colnames(activity), colnames(activity_effects))
  
  if (aligned == F) {
    warning(paste0("WARNING: Butterfly data and management strategy columns do not align/are not the same - Gt calculations will be incorrect. Iteration = ", i, ", 1=adult, 2=larvae"))
  }
  
  # find overlap
  overlap <- t(apply(activity, 1, function(row) row & activity_effects)) # NOTE: THIS ASSUMES THE COLUMNS ALIGN
  
  # sum rows to find Gt and store in output dataframe
  output[,i] <- rowSums(overlap)
  
  }
  
  return(output)
  
}
```

test calc Gt:
``` {r test-Gt}
calculate_Gta_Gtl(butterfly, man_strats[5,])[12,]
```

### Calculate Pt
Pta = Gta/(Ga*Ha)
Ptl = Gtl/(Gl*Hl)

``` {r calc-Pt}
calculate_Pta_Ptl <- function(data, strat) {
  
  G_H <- calculate_Ga_Gl_Ha_Hl(data)
  Ga <- G_H$Ga
  Gl <- G_H$Gl
  Ha <- G_H$Ha
  Hl <- G_H$Hl
  
  Gt <- calculate_Gta_Gtl(data, strat)
  Gta <- Gt$Gta
  Gtl <- Gt$Gtl
  
  # CALCULATE Pta
  Pta = Gta / (Ga * Ha)
  
  # replace infinite or NA values with 0
  Pta <- 
  ifelse(
    is.nan(Pta) | is.infinite(Pta) == T,  # test
    0, # if yes
    Pta  # if no
  )

  # CALCULATE Ptl
  Ptl = Gtl / (Gl * Hl)
  
  # replace infinite or NA values with 0
  Ptl <- 
  ifelse(
    is.nan(Ptl) | is.infinite(Ptl) == T,  # test
    0, # if yes
    Ptl  # if no
  )
  
  return(data.frame(Pta, Ptl))
}
```


## Calculate Ft (for a and l)
### Calculate A
$A$ = number of points of coincidence between the impact on and species’ use of forage plant families
### Calculate F
$F$ = points of coincidence between species habitat use and the location of its forage plants

``` {r calc-Fa}
calculate_Fa <- function(data, plant_locs) {
  
}
```