---
title: "Farmland Biodiversity Health Index - <br> Mammal Risk Scores"
author: "Elin Falla"
output:
  html_document: default
  pdf_document: default
  word_document: default
fontsize: 11pt
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction  
This script calculates risk scores associated with changes to farm management, for **butterfly** species reliant on farmland.  

**Note:** the script needs at least 2 species inputted to run.

The method is from **Butler, S. J. et al.** 'A cross-taxonomic index for quantifying the health of farmland biodiversity.' J. Appl. Ecol. 46, 1154-1162 (2009) https://doi.org/10.1111/j.1365-2664.2009.01709.x  

-----

The risk score is calculated as follows:  
$$ Risk = \frac{P_{ta} + F_{ta}}{R_a} + \frac{P_{tl} + F_{tl}}{R_l}$$

where subscript "l" refers to larvae and subscript "a" refers to adults. The risk score is hence the sum of the risk scores for the adult and larval stage for a particular species, where:  
$P_t$ = the risk score associated with reduced foraging activity potential 
$F_t$ = the risk score associated with reduction in forage plant availability
$R$ = the species’ reliance on farmland habitat  

$P_t$ is calculated as follows:  
$$P_t = \frac{G_t}{G*H}$$

$G_t$ = total number of generations of the species' given life stage (i.e. $G_{ta}$ for adults, $G_{tl}$ for larvae) active in the activity periods affected
$G$ = number of generations of the life stage across all activity periods (i.e. per year)
$H$ = number of habitat components used by the species' life stage  

-----  

$F_t$ is calculated as follows:  
$$N_t = \frac{A}{F}$$
$A$ = number of points of coincidence between impact on and species life stage's use of forage plant families  
$F$ = number of points of coincidence between species life stage's habitat use and the location of its forage plants  

## Set-up
### Load packages
```{r load-packages, message=FALSE}

rm(list = ls()) # clear env
library(readxl) # for reading in Excel sheets
library(tidyverse) # for data manipulation

```


### Source functions to calculate $P_t$ and $F_t$
Includes functions to calculate G, H, Gt, A, anf F, as well as Pt and Ft.  
``` {r source-Pt-Ft}
source("calc_Pt_Ft_functions.R")
```


### Read in required datasets  
Load datasets and replace all NAs with 0s.  
``` {r read-in-data}

# farm management strategies
man_strats <- read_xlsx("change.xlsx", 5) 

# butterfly requirements matrix data
butterfly <- read_xlsx("Butterfly_requirements.xlsx", skip = 3) 

# forage plant locations
plant_locs <- read_xlsx("forage_plant_location.xlsx", na="NA")

# butterfly forage plants
butterfly_plants <- read_xlsx("butterfly_forage_plants.xlsx")

# replace NAs with 0s in all datasets
man_strats <- man_strats %>% 
  mutate_all(~replace(., is.na(.), 0))

butterfly <- butterfly %>% 
  mutate_all(~replace(., is.na(.), 0))

plant_locs <- plant_locs %>% 
  mutate_all(~replace(., is.na(.), 0))

butterfly_plants <- butterfly_plants %>%
  mutate_all(~replace(., is.na(.), 0))
```

### Write function to create separate adult and larval datasets
``` {r sep-a-l}
separate_lifestages <- function(data) {
  
  ## Function to create separate datasets for adults and larvae
  
  if ("Lifestage" %in% colnames(data)) {  # this is the case for the butterfly forage plants matrix
    adults <- data[data$Lifestage == "a",]
    larvae <- data[data$Lifestage == "l",]
  }
  else {  # for the butterfly requirements matrix
    adults <- data %>%
    select(contains("Ad_"))
    
    # rename cols to   not include "Ad_"
    colnames(adults) <- gsub(pattern = "Ad_", replacement = "", colnames(adults)) 

    larvae <- data %>%
    select(contains("L_"))
    
    # rename cols to not include "L_"
    colnames(larvae) <- gsub(pattern = "L_", replacement = "", colnames(larvae))  
  }
  
  return(list(adults, larvae))
}
```


## Calculate $P_t$ (for adults, $P_{ta}$, and larvae, $P_{tl}$)

### Calculate $G$ and $H$
$G$ = number of generations across all activity periods (ie. number of generations per year)  
$H$ = number of habitat components used by the species 

### Calculate $G_t$
$G_t$ = number of generations of a species active in the activity periods affected

Here, 'strat' is a single management strategy - i.e. one row of the management strategy data frame.

number of each gens foraging habitats affected

### Calculate Pt
Pta = Gta/(Ga*Ha)
Ptl = Gtl/(Gl*Hl)

function = calculate_Pt()

## Calculate $F_t$ (for $a$ and $l$)

### Calculate $F$ and $A$
$F$ = points of coincidence between species habitat use and the location of its forage plants 
$A$ = number of points of coincidence between the impact on and species’ use of forage plant families

Butterflies have long tongues, so adults can forage on any plant in the habitat(s) they occupy. Therefore, Fa (F for adults) will just be the total number of forage plants found in the foraging habitat(s) of each butterfly species.

Larvae have much more specific plant forage families.


### Calculate $F_t$
function = calculate_Ft()

## Calculate risk score
``` {r calc-riskscore}
calc_butterfly_riskscore <- function(data, plants, plant_locs, strategies) {
  
  # initialise riskscores dataframe, rows are species, cols are strategies
  riskscores <- data.frame(matrix(NA, 
                                  nrow = nrow(data), 
                                  ncol = nrow(strategies) + 1)) # +1 to add in species names
  # put species names in first col
  riskscores[,1] <- data[,1]
  
  # name columns of riskscores according to strategies
  strat_names <- strategies %>% pull(agric_change)
  colnames(riskscores) <- c("species", strat_names)
  
  # extract reliance (R) values for adults and larvae from butterfly data
  R <- data %>% 
    select(contains("reliance"))
  Ra <- R %>%
    select(contains("Ad_"))
  Rl <- R %>%
    select(contains("L_"))
  
  # CALCULATE RISK SCORES per strategy and store in riskscores dataframe
  for (strat in 1:nrow(strategies)) {
    
    # separate adults and larval stages for both butterfly and plants data
    butterfly_data <- separate_lifestages(data)
    adult_data <- butterfly_data[[1]]
    larval_data <- butterfly_data[[2]]
    
    plants_data <- separate_lifestages(plants)
    adult_plants <- plants_data[[1]]
    larval_plants <- plants_data[[2]]
    
    # CALCULATE PT AND FT for both adults and larvae
    Pta <- calculate_Pt(adult_data, strategies[strat,])
    Ptl <- calculate_Pt(larval_data, strategies[strat,])
    
    Fta <- calculate_Ft(adult_data, adult_plants, plant_locs, strategies[strat,])
    Ftl <- calculate_Ft(larval_data, larval_plants, plant_locs, strategies[strat,])
    
    # CALCULATE RISK SCORE
    riskscores[,strat+1] <- ((Pta + Fta) / Ra) + ((Ptl + Ftl) / Rl)
  }
  
  # calculate total risk for each species by summing risk across all strategies (if more than one strategy)
  if (nrow(strategies) > 1) {
    riskscores$total_risk <- rowSums(riskscores[,-1])
  }
  
  return(riskscores)
}
  
```

### Test risk score
Test scores against Butler's risk scores to ensure method is correct.
``` {r test-riskscore}
risk <- calc_butterfly_riskscore(butterfly, butterfly_plants, plant_locs, man_strats)
risk
# 
# risk$species_latin <- c("Gonepteryx_rhamni", "Thecla_betulae", "Polygonum_c_album", "Polyommatus_icarus", "Thymelicus_lineola", 
#                         "Pyronia_tithonus", "Pieris_napi", "Celastrina_argiolus", "Ochlodes_sylvanus", "Pieris_brassicae",
#                         "Melanargia_galathea", "Euphydryas_aurina", "Maniola_jurtina", "Anthocharis_cardamines", "Aglais_io", 
#                         "Aphantopus_hyperantus", "Lycaena_phlaeas", "Coenonympha_pamphilus", "Thymelicus_sylvestris", "Aglais_urticae",
#                         "Pieris_rapae", "Pararge_aegeria", "Lasiommata_megera", "Satyrium_w_album")
# 
# risk_sorted <- risk[order(risk$species_latin),]
butler_risk <- c(2.19, 1.65, 3.81, 2.01, 1.42, 1.44, 2.10, 3.53, 1.67, 2.18, 3.03, 1.85, 2.02, 1.85, 2.10, 1.56, 2.34, 1.86, 1.65, 1.06, 2.42, 2.15, 2.10, 2.10)
butler_species <- c(
    "small_tortoiseshell",
    "orange_tip",
    "ringlet",
    "green-veined_white",
    "small_white",
    "holly_blue",
    "small_heath",
    "marsh_fritillary",
    "brimstone",
    "peacock",
    "wall_brown",
    "small_copper",
    "meadow_brown",
    "marbled_white",
    "large_skipper",
    "speckled_wood",
    "large_white",
    "comma",
    "common_blue",
    "gatekeeper",
    "white-letter_hairstreak",
    "brown_hairstreak",
    "essex_skipper",
    "small_skipper"
)
butler_data <- data.frame(butler_species, butler_risk)
butler_sorted <- butler_data[order(butler_data$butler_species),]

butler_sorted$my_risk <- round(risk$total_risk, 2)
butler_sorted$same <- butler_sorted$butler_risk == butler_sorted$my_risk
butler_sorted$diff <- butler_sorted$butler_risk - butler_sorted$my_risk
butler_sorted




```