---
title: "Farmland Biodiversity Health Index - <br> Mammal Risk Scores"
author: "Elin Falla"
output:
  html_document: default
  pdf_document: default
  word_document: default
fontsize: 11pt
---
```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction  
This script calculates risk scores associated with changes to farm management, for **mammal** species reliant on farmland.  

**Note:** the script needs at least 2 species inputted to run.

The method is from **Butler, S. J. et al.** 'A cross-taxonomic index for quantifying the health of farmland biodiversity.' J. Appl. Ecol. 46, 1154-1162 (2009) https://doi.org/10.1111/j.1365-2664.2009.01709.x  

-----  
 
The **risk score** is calculated as follows:  
$$ Risk = \frac{D_t + N_t}{R}$$
where:  
$D_t$ = the risk score associated with reduced food abundance or availability  
$N_t$ = the risk score associated with reduced breeding success  
$R$ = the species’ reliance on farmland habitat  

-----  

$D_t$ is calculated as follows:    
$$D_t = \frac{A}{D*F} + \frac{B}{F}$$
$D$ = total number of dietary components used by the species  
$F$ = total number of foraging habitat components used by the species  
$A$ = number of points of coincidence between the impact on and species’ use of dietary components  
$B$ = number of points of coincidence between the impact on and species’ use of foraging habitat components  

-----  

$N_t$ is calculated as follows:  
$$N_t = \frac{C1}{N} + \frac{C2}{N}$$
$N$ = number of nesting habitat components used by the species  
$C1$ = number of points of coincidence between potential impact on and the species’ use of nesting habitat components if impact is through reduced success in existing habitat  
$C2$ = number of points of coincidence between potential impact on and the species’ use of nesting habitat components if impact is through loss of habitat  

## 2. Set-up
### 2.1. Load packages
```{r load-packages, message=FALSE, purl=FALSE}

rm(list = ls()) # clear env
library(readxl) # for reading in Excel sheets
library(tidyverse) # for data manipulation

```

## 2.2. Load necessary functions

```riskscore_functions.R``` contains functions used by all risk score scripts. It contains:  

* **check_aligned()** - It is essential to the method used here to ensure that the columns of the management change dataset and species' requirements dataset align before calculating overlap parameters - i.e. parameters that work out the impact of a change on a specific resource for a species. This function checks this, and outputs a warning if the column names of the two inputted datasets are not identical.  

* **NA_to_zero()** - Replaces all NAs in a dataset with zeros.  

``` {r check-alignment, purl=FALSE}
source("riskscore_functions.R")
```

### 2.3. Read in required datasets  
Two input files are required to calculate mammal risk scores:  

* Species' requirements (variable name = "mammals") - requirements for survival per species, entered as binary data 

* Farm management changes (variable name = "mamm_man_strats") - binary data on how the farm management strategies affect the species' resources  

The files are read in and then all NA values are set to 0.  

Note that the farm management strategies are read in a separate chunk with ```purl = FALSE``` so if the file is sourced, a user can load their own management strategy dataset. It must however, follow the same format as the mamm_man_strats dataset.  

``` {r read-in-data}
# mammal requirements data
mammals <- read_xlsx("../data/Mammal_requirements.xlsx", skip = 2) 

# replace NAs with 0s
mammals <- NA_to_zero(mammals)
```

``` {r load-man-strats, purl=FALSE}
# farm management strategies
mamm_man_strats <- read_xlsx("../data/change.xlsx", sheet="mammals") 
mamm_man_strats <- mamm_man_strats[1:6,]  # just include strategies of interest (Butler strategies)

# replace NAs with 0s
mamm_man_strats <- NA_to_zero(mamm_man_strats)

```
## 3. Calculate $D_t$ 

### 3.1. Calculate $F$ and $D$  
$D$ = total number of dietary components used by the species  
$F$ = total number of foraging habitat components used by the species  

$D$ is not technically the number dietary components as aqueous invertebrates only exist in the riparian habitat. For example, for the polecat, by the official definition $D = 4$, but due to its feeding on aqueous invertebrates in the riparian habitat only, $D = 3.25$, giving 13 when multiplied by $F$ (which is 4 in this case). For this reason, and due to the fact that the dietary components are recorded per habitat, $D * F$ (the denominator of part of $D_t$) is calculated rather than $D$ alone. This is done by summing the dietary components of the data by row (species).  

$F$ is calculated by summing the foraging habitat components of the data by row (species), to give the total number of habitats used per species.  

``` {r calc-D}
calc_mamm_F_D <- function(data) {
  
  # calculate F
  F <- data %>%
    select(hab_start : hab_end,
           -c(hab_start, hab_end)) %>%
    rowSums(.)

  # calculate D * F
  D_times_F <- data %>%
    select(diet_start : diet_end,
           -c(diet_start, diet_end)) %>%
    rowSums(.)

  return(data.frame(D_times_F, F))
}
```

``` {r test-D, purl=FALSE}
all_species <- mammals$Species
data.frame(species = all_species,
           calc_mamm_F_D(mammals))
```

### 3.2. Calculate $A$ and $B$  
$A$ = number of points of coincidence between the impact on and species’ use of dietary components  
$B$ = number of points of coincidence between the impact on and species’ use of foraging habitat components  

In this function 'strat' is a single row from the management strategies database - i.e. one management strategy/change.  

The impact of the change on dietary and foraging habitat components ($A$ and $B$ respectively) is found by applying the '&' operation, which returns true only if both elements are non-zero. The management strategy is applied in this way row-wise to the mammals requirements data, and the row sums gives the total overlap, i.e. $A$ or $B$. This strategy (which is also used to calculate $C1$ and $C2$ later on) assumes that the columns in the two datasets are aligned. There is therefore a check that this is the case, with a warning returned if not.  

``` {r calc-A}
calc_mamm_A_B <- function(data, strat) {
  
  # subset diet and habitat columns from data
  diets_habs <- data %>%
    select(diet_start : hab_end)

  # subset relevant columns from strategy data
  diets_habs_effects <- strat %>%
    select(diet_start : hab_end)

  # get column number of last diet column
  num_hab_cols <- ncol(diets_habs %>% select(hab_start : hab_end))
  last_d_col <- ncol(diets_habs) - num_hab_cols
  
  # check requirements data and management change data align
  check_aligned(colnames(diets_habs), colnames(diets_habs_effects))
  
  # find overlap between diet + hab requirements and management change
  overlap <- data.frame(t(apply(diets_habs, 1, function(row) row & diets_habs_effects))) 
  names(overlap) <- names(diets_habs)
  
  # CALCULATE A (sum diet cols of overlap)
  A <- rowSums(overlap[,1:last_d_col])
  
  # CALCULATE B (sum habitat cols of overlap)
  B <- rowSums(overlap[,(last_d_col+1):ncol(diets_habs)])
  
  return(data.frame(A, B))
}

```

``` {r, purl=FALSE}
calc_mamm_A_B(mammals, mamm_man_strats[1,])
```



### 3.3. Calculate $D_t$  

$$D_t = \frac{A}{D*F} + \frac{B}{F}$$
This function calculates $D_t$ by calling the relevant previously-defined functions.  

``` {r calc-Dt}
calc_mamm_Dt <- function(data, strat) {
  
  A_B <- calc_mamm_A_B(data, strat)
  A <- A_B$A
  B <- A_B$B
  
   F_D <- calc_mamm_F_D(data)
   F <- F_D$F
   D_times_F <- F_D$D_times_F

  Dt <- A/(D_times_F) + B/F
  
  # replace infinite or NA values with 0
  Dt <- 
  ifelse(
    is.nan(Dt) | is.infinite(Dt) == T,  # test
    0, # if yes
    Dt  # if no
  )
  return(Dt)
}

```

``` {r, purl=FALSE}
all_species <- mammals$Species
out <- data.frame(species = all_species,
                  Dt = calc_mamm_Dt(mammals, mamm_man_strats[2,]))
out
```

## 4. Calculate $N_t$

### 4.1. Calculate $N$ 
$N$ = number of nesting habitat components used by the species 

$N$ is calculated by summing the nesting requirement columns for each row (species).  
``` {r calc-N}
calc_mamm_N <- function(data) {

  N <- data %>%
    select(nest_start : nest_end) %>%
    rowSums(.)

  return(N)
}

```

``` {r test-n, purl=FALSE}
all_species <- mammals$Species
data.frame(species = all_species,
           N = calc_mamm_N(mammals))

```

### 4.2. Calculate $C1$ and $C2$ 
$C1$ = number of points of coincidence between potential impact on and the species’ use of nesting habitat components if impact is through reduced success in existing habitat  
$C2$ = number of points of coincidence between potential impact on and the species’ use of nesting habitat components if impact is through loss of habitat  

Like for calculating $A$ and $B$ above, overlaps are calculated using the '&' operator - returns 1 if both values (i.e. the mammal nesting requirement and whether its affected by the management strategy) are non-zero.  

``` {r calc-C1C2}
calc_mamm_C1_C2 <- function(data, strat) {
  
  nesting <- data %>%
    select(nest_start : nest_end,
           -c(nest_start, nest_end))  # select nesting columns of mammal data
 
  nest_effects <- strat %>%
    select(nest_start : nest_end,
           -c(nest_start, nest_end))  # select nesting columns of strategy data
  
  # CALCULATE C1 - effects of reduced nesting success
  nest_success <- nest_effects %>%
    select(contains("_S"))  # select nesting success columns
      
  # Check columns of nest_success and nesting are the same (if not aligned, overlap calculation will not be correct)
  to_compare <- gsub(pattern = "_S", replacement = "", colnames(nest_success))  # remove "_S" from nesting_success colnames to compare
  check_aligned(to_compare, colnames(nesting)) # compare colnames
  
  # Calculate overlap
  C1_overlap <- t(apply(nesting, 1, function(row) row & nest_success))  
  C1 <- rowSums(C1_overlap)  # sum overlaps to get C1
  
  
  # CALCULATE C2 - effects of loss of nesting habitat
  nest_habs <- nest_effects %>%
    select(!contains("_S"))  # select habitat loss columns
   
  # Check columns of nest_habs and nesting are the same (if not aligned, overlap calculation will not be correct)
  check_aligned(colnames(nest_habs), colnames(nesting))
  
  # Calculate overlap
  C2_overlap <- t(apply(nesting, 1, function(row) row & nest_habs))  
  C2 <- rowSums(C2_overlap)  # sum overlaps to get C2
  
  return(data.frame(C1, C2))
}

```

### 4.3. Calculate $N_t$
$$N_t = \frac{C1}{N} + \frac{C2}{N}$$

``` {r calc-Nt}
calc_mamm_Nt <- function(data, strat) {
  
  N <- calc_mamm_N(data)
  
  C1_C2 <- calc_mamm_C1_C2(data, strat)
  C1 <- C1_C2$C1
  C2 <- C1_C2$C2

  Nt <- (C1/N) + (C2/N)
  
  # replace infinite or NA values with 0
  Nt <- 
    ifelse(
      is.nan(Nt) | is.infinite(Nt) == T,  # test
      0, # if yes
      Nt  # if no
    ) 
  
  return(Nt)
}

```

## 5. Calculate risk score
This function can take multiple strategies and outputs risk scores for each species for each strategy. It also (in the case that multiple strategies are inputted) calculates a **total risk score**, by summing the risk scores across all management strategies for each species.  

$$ Risk = \frac{D_t + N_t}{R}$$

``` {r calc-riskscore}
calc_mammal_riskscore <- function(data, strategies) {
  
  # initialise riskscores dataframe, rows are species, cols are strategies
  riskscores <- data.frame(matrix(NA, 
                                  nrow = nrow(data), 
                                  ncol = nrow(strategies) + 1)) # +1 to add in species names as first col
  # put species names in first col
  riskscores[,1] <- data[,1]
  
  # name columns of riskscores according to strategies
  strat_names <- strategies %>% pull(agric_change)
  colnames(riskscores) <- c("species", strat_names)
  
  # extract reliance (R) values from mammal data
  R <- data[,"Reliance"]  
  
  # CALCULATE RISK SCORES per strategy and store in riskscores dataframe
  for (strat in 1:nrow(strategies)) {
    
    Dt <- calc_mamm_Dt(data, strategies[strat,])
    Nt <- calc_mamm_Nt(data, strategies[strat,])
    
    # CALCULATE RISK SCORE
    riskscores[,strat+1] <- (Dt + Nt) / R
  }
  
  # calculate total risk for each species by summing risk across all strategies (if more than one strategy)
  if (nrow(strategies) > 1) {
    riskscores$total_risk <- rowSums(riskscores[,-1])
  }
  
  return(riskscores)
}
```

### 5.1. Calculate standardised risk score

In order to allow cross-taxonomic comparison of risk scores, they must be standardised. This is done by dividing the risk scores by the number of key resource requirements in the framework. This is equivalent to the number of overlap parameters - i.e. parameters that work out the impact of a change on a specific resource for a species. For mammals, this number is 4, as there are 4 such parameters: $A$, $B$, $C1$ and $C2$. See equations in Section 1 for more details.  

If only comparing risk scores within the taxon, this standardisation step is not required, and the ```calc_mammal_riskscore()``` function can be used instead.  

If one wants to calculate standardised risk scores from scratch (i.e. calculate risk scores then standardise them), the 'data' argument needs to be the species' requirements dataset, and 'strategies' needs to be the management strategies dataset. 'from_scratch' must be set to TRUE, and the 'risk_col' argument can be ignored.  

If one already has a dataset with risk scores in it that need to be standardised, 'data' needs to be said dataset, and 'risk_col' needs to be the name of the column with the risk scores in it. 'from_scratch' is FALSE by default and the 'strategies' argument can be ignored.

``` {r calc-standardised-riskscore}
calc_std_mammal_riskscore <- function(data, strategies=NA, from_scratch=FALSE, risk_col=NA) {
  
  # set number of resource requirements used by the taxa
  num_resource_reqs <- 4
  
  if (from_scratch == TRUE) {
    # calculate non-standardised risk scores
    risk_scores <- calc_mammal_riskscore(data, strategies)

  # divide all risk scores by number of resource requirements (minus the species name column)
    std_risk_scores <- risk_scores[,-1] / num_resource_reqs
    std_risk_scores <- cbind(risk_scores$species, std_risk_scores) # add back species name column
    names(std_risk_scores)[1] <- "species"
    
  }
  
  else {
    
    # extract risk score column
    risk_scores <- data %>%
      pull(risk_col)
    
    # create standard risk column
    std_risk_scores <- data %>%
      mutate(Std_risk = (risk_scores / num_resource_reqs))
  }
  
  return(std_risk_scores)
}
```

### 5.2. Test risk scores are correct
Calculate risk scores for the strategies used in Butler's paper in order to check they are the same.  

``` {r test-risk, purl=FALSE}
# Calculate risk scores
riskscores <- calc_mammal_riskscore(mammals, mamm_man_strats)

# Display risk scores
riskscores


butler_sp <- c(
"Yellow-necked mouse","Wood mouse","Water vole","Barbastelle","Roe deer","Red deer","Sika","Bank vole","Fallow deer","Serotine","Hedgehog","Chinese water deer","Brown hare","Otter","Badger","Harvest mouse","Field vole","Muntjac","House mouse","Common doormouse","Stoat","Weasel","Polecat","Bechsteins","Brandts","Daubentons","Whiskered","Natterers","Water shrew","Leisler","Noctule","Rabbit","Nathusius pipistrelle","Pipistrelle","Soprano pipistrelle","Brown long-eared","Grey long-eared bat","Brown rat","Greater horseshoe","Lesser horseshoe","Common shrew","Pygmy shrew","Common mole","Red fox"
)
butler_risk <- c(0.78,1.79,1.75,1.75,1.17,0.44,0.33,1.33,0.67,3.50,1.50,1.50,3.33,2.00,1.50,3.63,1.42,0.78,2.04,0.78,1.50,1.38,1.86,1.75,1.75,1.75,1.75,1.75,1.20,1.75,3.50,1.17,1.17,3.50,1.75,1.75,1.75,1.81,2.50,2.50,2.00,2.00,1.75,1.50
)
butler <- data.frame(butler_sp, butler_risk)
butler <- butler[order(butler$butler_sp),]
butler$my_risk <- round(riskscores$total_risk, 2)
butler$same <- butler$my_risk == butler$butler_risk
butler$diff <- butler$my_risk - butler$butler_risk

butler

print(paste("correct species =", nrow(butler[butler$same == T,]), "/", nrow(butler)))

```

